---
title: "Statistical analysis"
author: "David Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(multcomp) # for glht
library(broom)
library(cowplot)
theme_set(theme_cowplot())
```


```{r message = FALSE}
# read in Reilly's data
if (file.exists("data/life_history.csv")) {
  life_history <- read_csv("data/life_history.csv")
} else{
  life_history <- read_csv(url("https://raw.githubusercontent.com/reillyowencooper/daphnia_multigenerational_antibiotics/main/data/life_history.csv"))
  write_csv(life_history, "data/life_history.csv")
}

lh_df <-
  life_history %>% 
  rowwise() %>%
  mutate(total_offspring = sum(brood_one_numoffspring, brood_two_numoffspring, brood_three_numoffspring, na.rm = TRUE)) %>%
  mutate(genf = factor(gen),
         treatment = ifelse(treatment == "recovery", "control", treatment)) %>%
  mutate(sequence = substr(figure_labels, 1,2),
         sequence = ifelse(str_detect(sequence, "C"), "C", sequence),
         sequence = ifelse(str_detect(sequence, "A"), "A", sequence),
         sequence = factor(sequence)
         ) 
```


# Cell means model for Final Size

This model has a parameter for every single cell mean. That is, every sequence x generation combination for $i = 1,2,3, \ldots, 6$ sequences and $j = 0, 1, \ldots, 4$ generations. There are 20 combindations total (some combinations were not included in experimental design).

$$y_{ijk} = \tau_{ij} + \epsilon_{ijk}$$

* $y_{ijk}$ the final size of the $k^{th}$ daphnia in the $i^{th}$ sequence and $j^{th}$ generation
* $\tau_{ij}$ is the mean final size of daphnia in the $i^{th}$ sequence and $j^{th}$ generation
* $\epsilon_{ijk} \sim N(0,\sigma^2)$ is random error

```{r}
# create a col with cominations of sequence and generation
seq_gen_levels <- c(paste("C", ":", 0:4, sep = ""), 
                    paste("R1", ":", 1:4, sep = ""),
                    paste("R2", ":", 2:4, sep = ""),
                    paste("R3", ":", 3:4, sep = ""),
                    paste("R4", ":", 4, sep = ""),
                    paste("A", ":", 0:4, sep = ""))
# bod <- bod %>% mutate(seq_gen = paste(fsequence, ":", fgen, sep = ""),
#                       seq_gen = factor(seq_gen,
#                                        levels = seq_gen_levels))
lh_df <- lh_df %>% mutate(seq_gen = paste(sequence, ":", gen, sep = ""),
                      seq_gen = factor(seq_gen,
                                       levels = seq_gen_levels))


# fit model
# mod.size <- lm(final_size ~ seq_gen - 1, data = bod)
mod.size <- lm(final_size ~ seq_gen - 1, data = lh_df)
summary(mod.size)
```
```{r}
# tidy(mod.size) %>% mutate(term = str_remove(term, "seq_gen")) %>% knitr::kable()
# glance(mod.size)
```


# Model fit

```{r}
aug.size <-
  augment(mod.size, se_fit = TRUE) %>%
  mutate(width = qnorm(0.975)*sqrt(.se.fit),
         lwr = .fitted - width,
         upr = .fitted + width)
aug.size <- aug.size %>% inner_join(lh_df)

aug.size %>%
  ggplot() +
  geom_jitter(aes(x = gen, y = final_size), alpha = 0.3) +
  geom_point(aes(x = gen, y = .fitted), size = 2) +
  geom_errorbar(aes(x = gen, ymin = lwr, ymax = upr), width = 0) +
  facet_wrap(~sequence) +
  labs(y = "Final size in mm (95% CI)",
       x = "Generation")
```

Estimated cell means and 95% Wald confidence intervals unadjusted for family-wise error.

# Contrasts

```{r}
# get unique rows of model matrix
cells <- model.matrix(mod.size) %>% unique()
colnames(cells) <- colnames(cells) %>% str_remove(pattern = "seq_gen")

# name each row so we can reference it according to seq X gen combinations
# rownames(cells) <- names(actual_combos)

cell_rownames <- vector("character", length = nrow(cells))
for ( i in seq_along(cell_rownames)) {
  cell_rownames[i] <- colnames(cells)[cells[i,] == 1]
}

# clean up names
# colnames(cells) <- colnames(cells) %>% str_remove(pattern = "seq_gen")
rownames(cells) <- cell_rownames # colnames(cells)

# check that indexing is consistent with model coef order
# ((names(coef(mod.size)) %>% str_remove(pattern = "seq_gen")) == colnames(cells)) %>% sum() == length(coef(mod.size))
```

```{r}
# try using emmeans package
library(emmeans)
em.size <- emmeans(mod.size, "seq_gen")
em.sizedf <- data.frame(em.size)
```

```{r eval = FALSE}
# compare emmeans to observed cell means
# they are equal???
obs_mean <- bod %>% group_by(seq_gen) %>% summarise(cell_mean = mean(final_size, na.rm = TRUE)) 
inner_join(em.sizedf, obs_mean) %>% transmute(diff = near(emmean, cell_mean))
```

### Compare mean body size among sequences to control within each generation

```{r}
# put K into a list since ethat is what emmeans::contrast needs to specify linear hypothesis
contrast_list_in_gen <- list(cells["C:0",] - cells["A:0",],
           cells["C:1",] - cells["R1:1",],
           cells["C:1",] - cells["A:1",],
           cells["C:2",] - cells["R1:2",],
           cells["C:2",] - cells["R2:2",],
           cells["C:2",] - cells["A:2",],
           cells["C:3",] - cells["R1:3",],
           cells["C:3",] - cells["R2:3",],
           cells["C:3",] - cells["R3:3",],
           cells["C:3",] - cells["A:3",],
           cells["C:4",] - cells["R1:4",],
           cells["C:4",] - cells["R2:4",],
           cells["C:4",] - cells["R3:4",],
           cells["C:4",] - cells["R4:4",],
           cells["C:4",] - cells["A:4",])
hypothesis_names <- c("Generation 0: C - A",
                      "Generation 1: C - R1", "Generation 1: C - A",
                      "Generation 2: C - R1", "Generation 2: C - R2", "Generation 2: C - A",
                      "Generation 3: C - R1", "Generation 3: C - R2", "Generation 3: C - R3", "Generation 3: C - A",
                      "Generation 4: C - R1", "Generation 4: C - R2", "Generation 4: C - R3", "Generation 4: C - R4", 
                      "Generation 4: C - A")
names(contrast_list_in_gen) <- hypothesis_names

# get contrasts using mvt adjustment
LF_in_gen <- contrast(em.size, 
               contrast_list_in_gen,
               adjust = "mvt")
# save estimate, CI, and p.value
sum_in_gen <- summary(LF_in_gen) %>% data.frame()
contr_in_gen <- confint(LF_in_gen) %>% data.frame()
contr_in_gen <- contr_in_gen %>% add_column(p.value = sum_in_gen$p.value)

# shorten contrast name, make a col for generation
contr_in_gen <-
  contr_in_gen %>% 
  separate(contrast, into = c("generation", "contrast_name"), sep = ":", remove = FALSE) %>% 
  mutate(generation = str_remove(generation, "Generation "),
         generation = factor(generation))
```

```{r}
contr_in_gen %>%
  ggplot() +
  geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
  geom_point(aes(x = contrast, y = estimate, col = generation), size = 2) +
  geom_errorbar(aes(x = contrast, ymin = lower.CL, ymax = upper.CL, col = generation)) +
  scale_x_discrete(labels=contr_in_gen$contrast_name) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = "Comparison",
       y = "Difference in body size (mm)")
```

### Compare difference in body size over the course of the experiment

```{r}
# put K into a list since ethat is what emmeans::contrast needs to specify linear hypothesis
contrast_list_time <- list("C: gen 0 - gen 4" = cells["C:0",] - cells["C:4",],
                      "A: gen 0 - gen 4" = cells["A:0",] - cells["A:4",],
                      "gen 0 - gen 4: C - A" = (cells["C:0",] - cells["C:4",]) - (cells["A:0",] - cells["A:4",]))

LF_time <- contrast(em.size, 
               contrast_list_time,
               adjust = "mvt")
sum_time <- summary(LF_time) %>% data.frame()
contr_time <- confint(LF_time) %>% data.frame()
contr_time <- contr_time %>% add_column(p.value = sum_time$p.value)
```

```{r}
contr_time %>%
  ggplot() +
   geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
  geom_point(aes(x = contrast, y = estimate), size = 2) + 
  geom_errorbar(aes(x = contrast, ymin = lower.CL, ymax = upper.CL), width = 0) +
  labs(y = "difference in body size (mm)")
```
### Compare the mean body size at first recovery time by number of generations exposed to antibiotics

```{r}
contrast_list_first_recovery <- 
  list("R1 - R2" = cells["R1:1",] - cells["R2:2",],
       "R1 - R3" = cells["R1:1",] - cells["R3:3",],
       "R1 - R4" = cells["R1:1",] - cells["R4:4",],
       "R2 - R3" = cells["R2:2",] - cells["R3:3",],
       "R2 - R4" = cells["R2:2",] - cells["R4:4",],
       "R3 - R4" = cells["R3:3",] - cells["R4:4",])

LF_first_recovery <- contrast(em.size,
                              contrast_list_first_recovery,
                              adjust = "mvt")
sum_first_recovery <- summary(LF_first_recovery) %>% data.frame()
first_recovery <- confint(LF_first_recovery) %>% data.frame()

first_recovery <- first_recovery %>% add_column(p.value = sum_first_recovery$p.value)

first_recovery %>%
  ggplot() +
   geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
  geom_point(aes(x = contrast, y = estimate), size = 2) + 
  geom_errorbar(aes(x = contrast, ymin = lower.CL, ymax = upper.CL), width = 0) +
  labs(y = "difference in body size (mm)")
```


### Model Checking

```{r}
# model checking
plot(mod.size)
```

Everything looks good.

# Cumulative reproduction analysis

```{r }
mod.repro <- lm(total_offspring ~ seq_gen - 1, data = lh_df)
summary(mod.repro)
```

```{r}
plot(mod.repro)
```

Definitely some heteroscedasticity. The variance seems to increase with the mean. Poisson would be better.

```{r}
aug.repro <- augment(mod.repro, se_fit = TRUE) %>% 
  mutate(width = qnorm(0.975)*sqrt(.se.fit),
         lwr = .fitted - width,
         upr = .fitted + width)
aug.repro <- aug.repro %>% inner_join(lh_df)

aug.repro %>%
  ggplot() +
  geom_jitter(aes(x = gen, y = total_offspring), alpha = 0.3) +
  geom_point(aes(x = gen, y = .fitted), size = 2) +
  geom_errorbar(aes(x = gen, ymin = lwr, ymax = upr), width = 0) +
  facet_wrap(~sequence) +
  labs(title = "Model assuming normally distributed responses",
       y = "Cumulative offspring (95% CI)",
       x = "Generation")
```

The confidence intervals (95% Wald CI) are waaaay too small. This means the probability of rejecting null hypotheses will be much bigger than intended. Not a great model. Really think I should use count data model.

```{r}
mean_var <-
  lh_df %>%
  group_by(sequence, gen) %>%
  transmute(mean_offspring = mean(total_offspring),
         var_offspring = var(total_offspring)) %>%
  slice(1)

varmean.nb <- lm(var_offspring ~ I(mean_offspring^2) - 1, data = mean_var)
varmean.qp <- lm(var_offspring ~ mean_offspring - 1, data = mean_var)
nd_offspring <- data.frame(mean_offspring = seq(0, 85, by = 1))
nd_offspring$pred_nb <- predict(varmean.nb, newdata = nd_offspring)
nd_offspring$pred_qp <- predict(varmean.qp, newdata = nd_offspring)

ggplot() +
  geom_point(data = mean_var, mapping = aes(x = mean_offspring, y = var_offspring)) +
  geom_line(data = nd_offspring, mapping = aes(x = mean_offspring, y = pred_nb)) +
  geom_line(data = nd_offspring, mapping = aes(x = mean_offspring, y = pred_qp)) +
  geom_abline(slope = 1) +
  ylim(0,85) + xlim(0,85)
```

Here, I've plotted the sequence x generation specific mean and variance in offspring along with a x = y line. The variance is always greater than the mean so this suggests I will need to use negative binomial or quasipoisson regression. Negative binomial regression would imply variance is a function of the square of the mean and quasipoisson allows variance to be a linear function of the mean. Either one seems ok. I will need to actually fit these models and compare them.

I think there are two possible strategies here:

1. Fit count data models. May need to fit binary response model (reproduce or not reproduce) and another model for number of offspring (given that you reproduced, how many babies did you have?)
2. Calculate the per-capita growth rate (r) for the daphnia, which would take both survival and reproduction into account, and model how r depends on the treatment and generation.

I think the 2^nd^ option makes the most sense if you want to make statements about how the treatments impact fitness, since it is a direct measurement of fitness. Although, I'm not sure if it makes sense to calculate r for individual daphnia. But, i think alaina says that she has done it and has a spreadsheet to do so.

Just how many Daphnia didn't have offspring?

```{r}
# individuals with no offspring
none_df <- lh_df %>%
    group_by(seq_gen) %>% filter(total_offspring == 0) %>%
  dplyr::select(gen, treatment, final_survival_day, final_size,
                brood_one_numoffspring, brood_two_numoffspring, brood_three_numoffspring, total_offspring)
```

number of daphnia that had no babies
```{r}
lh_df %>%
  group_by(seq_gen) %>% filter(total_offspring == 0) %>%
  summarise(no_babies = n()) %>% knitr::kable()
```

number of daphnia that survived to 21 days and had no babies
```{r}
lh_df %>%
  group_by(seq_gen) %>% filter(total_offspring == 0, !is.na(final_size)) %>%
  summarise(no_babies = n()) %>% knitr::kable()

```

Total offspring by final survival day

```{r}
lh_df %>%
  ggplot() +
  geom_jitter(aes(x = final_survival_day, y =total_offspring), alpha = 0.5) +
  geom_hline(yintercept = 0, col = "red") +
  facet_grid(sequence~gen)
```

Whats up with the A:4 individual that had 20 babies and died at 1 day old?

```{r}
lh_df %>% filter(sequence == "A", total_offspring > 10, final_survival_day < 10) %>%
    dplyr::select(gen, treatment, final_survival_day, final_size,
                brood_one_numoffspring, brood_two_numoffspring, brood_three_numoffspring, total_offspring) %>%
  knitr::kable()
```

Yeah, there is definitily a data entry error here. It looks like there are two records where the day of first brood is greater than the final survival day. They were probably swapped on accident.

```{r}
lh_df %>% filter(brood_one_day > final_survival_day) %>% 
  dplyr::select(gen, treatment, individual, final_survival_day, brood_one_day) %>% 
  knitr::kable()
```


```{r echo = FALSE}
knitr::knit_exit()
```

# contrasts using multcomp::glht
```{r}
# function to convert glht object to data.frame
# https://gist.github.com/cheuerde/3acc1879dc397a1adfb0
# Claas Heuer, September 2015
glht.table <- function(x) {

# I took this from somewehre, but cant remember the source (probably SO))
  pq <- summary(x)$test
  mtests <- cbind(pq$coefficients, pq$sigma, pq$tstat, pq$pvalues)
  error <- attr(pq$pvalues, "error")
  pname <- switch(x$alternativ, less = paste("Pr(<", ifelse(x$df ==0, "z", "t"), ")", sep = ""), 
  greater = paste("Pr(>", ifelse(x$df == 0, "z", "t"), ")", sep = ""), two.sided = paste("Pr(>|",ifelse(x$df == 0, "z", "t"), "|)", sep = ""))
  colnames(mtests) <- c("Estimate", "Std. Error", ifelse(x$df ==0, "z value", "t value"), pname)
  return(mtests)

}

# create contrast matrix: C - (other sequence) within all generations
K <- rbind(cells["C:0",] - cells["A:0",],
           cells["C:1",] - cells["R1:1",],
           cells["C:1",] - cells["A:1",],
           cells["C:2",] - cells["R1:2",],
           cells["C:2",] - cells["R2:2",],
           cells["C:2",] - cells["A:2",],
           cells["C:3",] - cells["R1:3",],
           cells["C:3",] - cells["R2:3",],
           cells["C:3",] - cells["R3:3",],
           cells["C:3",] - cells["A:3",],
           cells["C:4",] - cells["R1:4",],
           cells["C:4",] - cells["R2:4",],
           cells["C:4",] - cells["R3:4",],
           cells["C:4",] - cells["R4:4",],
           cells["C:4",] - cells["A:4",]
           )
# Labels for hypothesis test
hypothesis_names <- c("Generation 0: C - A",
                      "Generation 1: C - R1", "Generation 1: C - A",
                      "Generation 2: C - R1", "Generation 2: C - R2", "Generation 2: C - A",
                      "Generation 3: C - R1", "Generation 3: C - R2", "Generation 3: C - R3", "Generation 3: C - A",
                      "Generation 4: C - R1", "Generation 4: C - R2", "Generation 4: C - R3", "Generation 4: C - R4", 
                      "Generation 4: C - A")

rownames(K) <- hypothesis_names

# run all hypothesis tests
size.ht <- glht(mod.size, linfct = K)
```

```{r}
ci.size <- confint(size.ht)$confint %>% data.frame()
p.size <- glht.table(size.ht) %>% data.frame() 

hypoth.df <- inner_join(p.size, ci.size, by = "Estimate") %>% mutate(test = hypothesis_names) %>% relocate(test)
hypoth.df %>%  knitr::kable()
```

```{r}
hypoth.df %>%
  ggplot(aes(x = test)) +
  geom_point(aes(y = Estimate)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr)) +
  geom_hline(yintercept = 0, col = "red", linetype = "dashed") +
#  scale_x_discrete(limits = rev(levels(seq_gen_levels))) +
  coord_flip() +
  labs(title = "Difference in final body size within each generation",
       subtitle = "Control is reference",
       x = "",
       y = "Estimated difference (95% family-wise CI)") +
  theme_bw()
```

```{r eval = FALSE, echo = FALSE}
# get unique rows of model matrix
# includes all seq:gen (even those missing in experiment)
cells <- model.matrix(mod.size) %>% unique()

# get names of actual seq X gen combinations
all_combos <- apply(cells, 2, sum)
actual_combos <- all_combos[all_combos == 1]

# name each row so we can reference it according to seq X gen combinations
rownames(cells) <- names(actual_combos)

# notation: sequence.generation
# Control - (not control) for each generation
K <-
rbind(cells["fsequenceC:fgen0",] - cells["fsequenceA:fgen0",], # C.0 - A.0
      cells["fsequenceC:fgen1",] - cells["fsequenceR1:fgen1",], # C.1 - R1.1
      cells["fsequenceC:fgen1",] - cells["fsequenceA:fgen1",], # C.1 - A.1
      cells["fsequenceC:fgen2",] - cells["fsequenceR1:fgen2",], # C.2 - R1.2
      cells["fsequenceC:fgen2",] - cells["fsequenceR2:fgen2",], # C.2 - R2.2
      cells["fsequenceC:fgen2",] - cells["fsequenceA:fgen2",],  # C.2 - A.2
      cells["fsequenceC:fgen3",] - cells["fsequenceR1:fgen3",], # C.3 - R1.3
      cells["fsequenceC:fgen3",] - cells["fsequenceR2:fgen3",], # C.3 - R2.3
      cells["fsequenceC:fgen3",] - cells["fsequenceR3:fgen3",], # C.3 - R3.3
      cells["fsequenceC:fgen3",] - cells["fsequenceA:fgen3",],  # C.3 - A.3
      cells["fsequenceC:fgen4",] - cells["fsequenceR1:fgen4",], # C.4 - R1.4
      cells["fsequenceC:fgen4",] - cells["fsequenceR2:fgen4",], # C.4 - R2.4
      cells["fsequenceC:fgen4",] - cells["fsequenceR3:fgen4",], # C.4 - R3.4
      cells["fsequenceC:fgen4",] - cells["fsequenceR4:fgen4",], # C.4 - R4.4
      cells["fsequenceC:fgen4",] - cells["fsequenceA:fgen4",])  # C.4 - A.4

# remove cols that don't actually exist in experiment
K <- K[, colnames(K) %in% names(actual_combos)]

# check that K is correct
apply(K, 1, sum) == 0 # contrasts should sum to zero

size.ht <- glht(mod.size, linfct = K)
```

```{r eval = FALSE}
# note there is some missing data
bod %>% group_by(sequence, gen) %>% summarise(n())

# missing response
bod %>% group_by(sequence, gen) %>% filter(is.na(final_size)) %>% summarise(n())

# check out number of obs for each carryover parameter
bod %>% group_by(carryover) %>% summarise(n())
```

```{r}
bod %>% ggplot(aes(x = fgen, y = final_size)) + geom_jitter(alpha=0.5) + facet_wrap(~sequence)
```


```{r}
# fit fixed effects model
mod1 <- lm(final_size ~ #-1 +
            ftreatment*fgen,
          data = bod)
bod$carryover <- factor(bod$carryover)
contrasts(bod$carryover) <- contr.sum
# contrasts(bod$ftreatment) <- contr.sum

mod2 <- lm(final_size ~ -1 +
            fcarryover +
            ftreatment*fgen, 
          data = bod)

lm(final_size ~ fcarryover + ftreatment*fgen, data = bod)
model.matrix(~ ftreatment*fgen + carryover, data = bod) %>% mgcv::uniquecombs() %>% View()

bod <- drop_na(bod) # WRONG. drops all gen == 0, because carryover == NA
bod <- bod %>% filter(!is.na(final_size)) # doesn't drop NA in carryover

X <- model.matrix(~ ftreatment*fgen + carryover - 1, data = bod)
#X %>% mgcv::uniquecombs() %>% View()
lsbeta <- MASS::ginv(t(X) %*% X) %*% t(X) %*% # matrix operations drop NAs, so all gen0 is lost since carryover NA
  as.matrix(bod[,"final_size",drop = FALSE])
names(lsbeta) <- colnames(X)
lsbeta["ftreatmentcontrol"] + lsbeta["fgen1"]
# summary(mod1)
summary(mod2)
# model.matrix(mod2) %>% as.tibble() %>% distinct() %>% View()
# car::Anova(mod1, type = 3)
car::Anova(mod2, type = 3)
```

Is it appropriate to set gen0 carryover as C for both Control and ab sequences, since they each come from a "control" source population?

```{r}
aovmod <- aov(final_size ~ genf*treatment, data = bod)
summary(aovmod)
coef(aovmod)
TukeyHSD(aovmod)
```


```{r}
# residual diagnostics
plot(mod2)
```

```{r}
aug <- augment(mod2)
inner_join(bod, aug) %>%
  ggplot() +
  geom_jitter(aes(x = fgen, y = final_size), alpha = 0.5) +
  geom_point(aes(x = fgen, y = .fitted), size = 2, col = "red") + facet_wrap(~fsequence)
```

```{r}
# cell means model
mod.cm <- lm(final_size ~ fsequence:fgen - 1, data = bod)
summary(mod.cm)
```

Estimates a parameter for every sequence x generation group (20 total).

```{r message = FALSE, eval = FALSE, echo = FALSE}
# read in and format data for analysis
if (file.exists("data/bod.csv")){
  bod <- read.csv("data/bod.csv") %>%
    mutate(fgen = factor(gen),
           fsequence = factor(sequence, levels = c("C", "R1", "R2", "R3", "R4", "A")),
           ftreatment = factor(treatment, levels = c("control","antibiotic")),
           fcarryover = factor(carryover, levels = c("C", "A")))
} else { # if file not found read in raw data and process
  # read in Reilly's data
life_history <- read_csv(url("https://raw.githubusercontent.com/reillyowencooper/daphnia_multigenerational_antibiotics/main/data/life_history.csv"))

# select body size data
bod <- 
  life_history %>% 
  select(gen, treatment, individual, final_size, gen_in_control, gen_in_antibiotics, figure_labels) %>%
  mutate(genf = factor(gen),
         treatment = ifelse(treatment == "recovery", "control", treatment))

# create "sequence" to ID each unique sequence
bod <- 
  bod %>%
  mutate(sequence = substr(figure_labels, 1,2),
         sequence = ifelse(str_detect(sequence, "C"), "C", sequence),
         sequence = ifelse(str_detect(sequence, "A"), "A", sequence),
         sequence = factor(sequence)
         ) 
# create "carryover" covariate to specify the previous treatment
bod <-
  bod %>% 
  group_by(gen, sequence) %>%
  # mutate(carryover = case_when( (sequence == "A" & gen > 0) ~ "A",
  #                               (sequence == "C" & gen > 0) ~ "C",
  #                               (str_extract(sequence, "\\d") == gen) ~ "A",
  #                               (str_extract(sequence, "\\d") != gen) ~ "C"
  #                              ))
  mutate(carryover = case_when( (sequence == "A" & gen > 0) ~ "A",
                                (sequence == "C" & gen > 0) ~ "C",
                                (str_extract(sequence, "\\d") == gen) ~ "A",
                                (str_extract(sequence, "\\d") != gen) ~ "C",
                                gen == 0 ~ "C"
                               ))
# save csv
write_csv(bod, "data/bod.csv")

# format bod for analysis in SAS
# bod <- drop_na(bod)
bod %>% 
  select(gen, treatment, individual, sequence, 
         carryover, final_size) %>%
  # mutate(sequence = case_when(sequence == "C" ~ "C",
  #                             sequence == "R1" ~ "Rone",
  #                             sequence == "R2" ~ "Rtwo",
  #                             sequence == "R3" ~ "Rthr",
  #                             sequence == "R4" ~ "Rfou",
  #                             sequence == "A" ~ "A")) %>%
  mutate(sequence = case_when(sequence == "C" ~ 1,
                              sequence == "R1" ~ 2,
                              sequence == "R2" ~ 3,
                              sequence == "R3" ~ 4,
                              sequence == "R4" ~ 5,
                              sequence == "A" ~ 6)) %>%
  write_csv("data/bod_sas.csv")
}
```
